Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-20 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex

We do not expect any fundamental complications to arise if the poloidal and toroidal magnetic drifts
were retained in the kinetic equation. 
However, we have several reasons for neglecting these magnetic drifts
 in this manuscript.
\begin{enumerate}
\item The ratio of these magnetic drifts to the parallel streaming term we keep is $\sim \rho_* = \rho / L$
 (the gyroradius to scale length), so neglecting the poloidal and toroidal magnetic drifts is
equivalent to considering the $\rho_* \to 0$ limit.
Since our primary interest in this paper is to compare the various options for the electric field
terms and collision operator, it is reasonable to take this $\rho_* \to 0$ limit for simplicity.
\item
Retaining these drifts would make the transport matrix (and figures 1, 2, 4, and 4)
depend on $\rho_*$ in addition to $\nu'$ and $E_*$. 
The dependence of the matrix elements on $\nu'$ and $E_*$ is already  complicated
and interesting to discuss in the $\rho_* \to 0$ limit,
without the extra complexity of finite-$\rho_*$ effects.
%\item Including these magnetic drift terms would also introduce dependence on the radial
%derivatives of the magnetic field
\item The poloidal and toroidal magnetic drifts would add extra complicating
terms to the analysis of moments of the kinetic equation in section III.
%\item If the poloidal and toroidal magnetic drifts are retained without the radial magnetic drift, ????
%\item The poloidal and magnetic drift terms will take additional time to implement in the code, 
%and the 
\end{enumerate}

We have added some text to section II (between (15)-(16)) and a sentence to section V (between \todo{(35)-(36)}) to clarify these issues.

@@

First, consider the case $E_r=0$, so there is no difference between the trajectory models.
Consider the homogeneous solutions of the kinetic equation, i.e. the null space of the linear system
we solve if no sources or constraints are included. 
When the linearized Fokker-Planck operator is used, 
there are 2 linearly independent homogeneous solutions:
a Maxwellian flux function,
and a Maxwellian flux function times $v^2$.
However, if the pure pitch-angle scattering operator or momentum-conserving model
collision operator are used, there are $N_x$ independent homogeneous solutions,
since any function of $v$ (independent of $\theta$,$\zeta$, and $\xi$)
is in the null space.

\todo{Now consider the case of nonzero $E_r$...}

\begin{referee}
Concerning the source/sink term, I think that the usage of such an artificial term is relevant if
the effect of the term is small. Can you show the effect of the source/sink term is small or not in
some cases shown in Section VI, by comparing $\left< \int d^3v\, S(\psi,v) \right> \Delta t$ and $n(\psi)$, for example, where
$\Delta t$ is some characteristic time scale such as collision time or transport time scale.
\end{referee}

The sources normalized to the transport timescale are now plotted in figure 3.
As discussed in the new text accompanying this figure, 
the sources normalized by the ion collision time are only a factor $1/0.4=2.5$ larger than the values plotted,
so the source terms remain much smaller than the collision operator in the kinetic equation.

### 2014-02-20 doc/sfincsPaper/sfincsPaper.tex
Added sources to fig 3 in manuscript to address referee's question about source magnitude

\changed{
We close this section by noting another an important difference between the trajectory models
related to the parallel momentum equations.
Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
current vanishes for all values of $d \Phi_0/d\psi$
if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
is known as intrinsic ambipolarity.
Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
when the $E_r$ terms are retained in the full trajectory
drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
moment. For the full trajectories, 
(\ref{eq:fullViscosity}) vanishes in this spatial average, leaving
\begin{equation}
\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
\label{eq:noCurrent}
\end{equation}
meaning there is no radial current. However, for the DKES and partial trajectory models, the
spatial average does not annihilate the $d\Phi_0/d\psi$ terms,
leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
and therefore the radial current is generally nonzero.  
Consequently, the full trajectory model is the only one of the models that preserves
intrinsic ambipolarity in quasisymmetry for $E_r \ne 0$.
Notice that when the full trajectory model is applied in quasisymmetry,
intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
so the net heat source vanishes for any radial electric field.
}

@@

\begin{figure}[h!]
% Figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
% SFINCS run on hopper and edison in 
% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
\includegraphics{m20131212_02_PlotSFINCS2SpeciesEScanForPaper.eps}
\caption{(Color online)
Fluxes and flows computed for a scenario of steep pressure gradient near the edge of W7-X.
Magenta dotted line is the ambipolar $E_r$, effectively identical for the three
trajectory models.
\label{fig:2species}}
\end{figure}

>>>>

\changed{
We close this section by noting another an important difference between the trajectory models
related to the parallel momentum equations.
Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
current vanishes for all values of $d \Phi_0/d\psi$
if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
is known as intrinsic ambipolarity.
Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
when the $E_r$ terms are retained in the full trajectory
drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
moment. For the full trajectories, 
(\ref{eq:fullViscosity}) vanishes in this spatial average, leaving
\begin{equation}
\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
\label{eq:noCurrent}
\end{equation}
meaning there is no radial current. However, for the DKES and partial trajectory models, the
spatial average does not annihilate the $d\Phi_0/d\psi$ term,
leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
and therefore the radial current is generally nonzero.  
Consequently, the full trajectory model is the only one of the models that preserves
intrinsic ambipolarity in quasisymmetry for $E_r \ne 0$.
Notice that when the full trajectory model is applied in quasisymmetry,
intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
so the net heat source vanishes for any radial electric field.
}

@@

\begin{figure}[h!]
% Figure generated using m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.m
% ----------------------------------------
% Previous version of figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
% SFINCS run on hopper and edison in 
% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
\includegraphics[height=7.5in]{m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.eps}
%\includegraphics{m20131212_02_PlotSFINCS2SpeciesEScanForPaper.eps}
\caption{(Color online)
Fluxes (A)-(B), flow (C), and bootstrap current (D)
computed for a scenario of steep pressure gradient near the edge of W7-X.
Magenta dotted line is the ambipolar $E_r$, effectively identical for the three
trajectory models.
\changed{
The ion particle and heat sources in (D)-(E) are normalized
by a gyro-Bohm transport time $t_{gB}$.}
\label{fig:2species}}
\end{figure}

\changed{
Figures (\ref{fig:2species}).E-F illustrate the ion particle and heat sources computed as part of the calculation.
As expected, the particle and heat sources are zero for the DKES model,
and for the full trajectory model, 
the particle source is always zero and the heat source vanishes at the ambipolar $E_r$.
Electron sources are negligible.
The plots show $S_{H\mathrm{p}}$ and $S_{H\mathrm{h}}$ from (\ref{eq:sources}) normalized to a gyro-Bohm transport time scale
$t_{gB} = L^2/D_{gB}$ with $D_{gB} = (\rho_i / L) cT/(eB_0)$, thereby
roughly normalizing the numerical sources to the scale of real physical sources arising from 
the divergence of the turbulent and neoclassical fluxes.
For this comparison we choose $L = -n / (dn/dr)$ to be the density scale length.
For the range of electric fields considered, the numerical sources are small on this transport time scale,
giving confidence in the model.
For the parameters considered, $\nuii t_{gB} = 0.4$, 
so dividing the values in figures (\ref{fig:2species}).E-F by this factor,
the source terms evidently remain much smaller
than the collision term in the kinetic equation for this calculation.
}
